# Markov Chains

```{r}
#| echo=FALSE
library(tibble)
library(ggplot2)
library(grid)        # for arrow()
library(ggdiagram)
```

<!-- In this practical you will learn a number of techniques in R that you wie_r_c <- connect(
  from = s["Rainy"], to = s["Cloudy"],
  label = lbl_04, label_sloped = FALSE, resect = 2, color = "black",
  curvature = 0.3
)
e_r_s <- connect(
  from = s["Rainy"], to = s["Sunny"],
  label = lbl_03, label_sloped = FALSE, resect = 2, color = "black",
  curvature = -0.4
)simulate simplified games of Monopoly (<https://en.wikipedia.org/wiki/Monopoly_(game)>). In addition, there are also many tutorials and guides on the Web describing how to produce computer simulations for Monopoly. You are welcome to read and use these examples to inspire your work. -->

We will now look at a **Markov Chain**. We have not covered it during lectures but based on the basic principles we have covered we will be able to use it for simulations.

Any random process is known to have the *Markov property* (a Markov process) if the probability of going to the next state depends only on the current state and not on the past states. A Markov process is **memoryless property** in that it does not store any property or memory of its past states.

If a Markov process operates within a specific (finite) set of states, it is called a ***Markov Chain***.

A Markov Chain is defined by three properties:

1.  A state space: a set of values or states in which a process could exist

2.  A transition matrix: defines the probability of moving from one state to another state

3.  A current state probability distribution: defines the probability of being in any one of the states at the start of the process

Consider the following example where we have two states describing the weather on any particular day: (i) Sunny and (ii) Rainy. Each arrow denotes the probability of going from one state to itself or another over the course of a day. For example, if it is currently sunny, the probability of it raining the next day is 0.6. Conversely, if it is raining, the probability that it will become sunny the next day is 0.7 and 0.3 that it will continue raining.

```{r tstate}
#| echo=FALSE

# ===== CONFIGURATION VARIABLES =====
my_font <- "Fira Code"
my_font_size <- 18
my_point_size <- 2
# my_colors <- viridis::viridis(2, begin = .25, end = .5)
my_colors <- c("#440154FF", "#2C728EFF", "#5DC863FF")


# Define node style and large label style
node_style1 <- ob_style(
  fill = my_colors[1],
  color = "white"
)

node_style2 <- ob_style(
  fill = my_colors[2],
  color = "white"
)

large_node_label <- redefault(
  ob_label,
  color = "white",
  fill = NA,
  size = 20,
  family = my_font
)
# Create nodes directly with custom styling
sunny_node <- ob_circle(
  x = -2,
  style = node_style1,
  label = large_node_label("Sunny")
)

rainy_node <- ob_ellipse(
  x = 2,
  style = node_style2,
  label = large_node_label("Rainy")
)

# Position the nodes
sunny_node <- place(sunny_node, sunny_node@bounding_box, "left", sep = 0.5)
rainy_node <- place(rainy_node, sunny_node@bounding_box, "right", sep = 3)

# ---- cross edges (use your connect + on-path labels) ----
lbl06 <- ob_label("0.6", angle = 0, position = 0.50, size = 18, color = my_colors[1])
lbl07 <- ob_label("0.7", angle = 0, position = 0.50, size = 18, color = my_colors[2])

e_sr <- connect(
  from = sunny_node, to = rainy_node,
  label = lbl06, label_sloped = FALSE, resect = 1,
  color = my_colors[1],
  to_offset = ob_point(0,  3)
)
e_rs <- connect(
  from = rainy_node, to = sunny_node,
  label = lbl07, label_sloped = FALSE, resect = 2, color = my_colors[2],
  to_offset = ob_point(0, -3)
)
e_ss <- connect(
  from = sunny_node, to = sunny_node,
  label = ob_label("0.4", angle = 0, position = 0.50, size = 18, color = my_colors[3]),
  label_sloped = FALSE, resect = 2, color = my_colors[3],
  from_offset = ob_point(-2, 3), to_offset = ob_point(2, 3)
)
e_rr <- connect(
  from = rainy_node, to = rainy_node,
  label = ob_label("0.3", angle = 0, position = 0.50, size = 18, color = my_colors[3]),
  label_sloped = FALSE, resect = 2, color = my_colors[3],
  from_offset = ob_point(-3, -2), to_offset = ob_point(3, -3)
)

# ---- base diagram with larger text ----
p <- ggdiagram(font_family = my_font, font_size = 28) + sunny_node + rainy_node
for (ed in list(e_sr, e_rs, e_ss, e_rr)) {
  p <- p + ed
}

print(p)

```

The transition matrix can be written as the following in R:

```{r}
transitionMatrix = matrix(c(0.4, 0.6, 0.7, 0.3), nrow=2, ncol=2, byrow=TRUE)
print(transitionMatrix)
```

which creates a 2 x 2 matrix consisting of the transition probabilities shown in the diagram.

Suppose I want to simulate a sequence of 30 days and the weather patterns over those days. Assuming that on day 0 it is currently sunny, I can do the following:

```{r}
# initial state - it is [1] sunny or [2] rainy
state <- 1
weather_sequence <- rep(0, 30) # vector to store simulated values
for (day in 1:30) { # simulate for 30 days
  pr <- transitionMatrix[state, ] # select the row of transition probabilities

  # sample [1] or [2] based on the probs pr
  state = sample(c(1, 2), size = 1, prob = pr)
  weather_sequence[day] <- state # store the sampled state
}

# print the simulated weather sequence
print(weather_sequence)
```

::: {.infobox}
## Exercise MC {#mc-exercise}

> Can you extend this example to a three-state model?

```{r fstate}
#| echo=FALSE,
#| warning=FALSE,
#| fig.bg = 'transparent'

# ===== CONFIGURATION VARIABLES =====
my_font <- "Fira Code"
font_size <- 18
label_font_size <- 20
node_label_size <- 0.8 # Font size for node labels (Cloudy, Sunny, Rainy)

# Node configuration
node_radius <- 15
node_separation <- 50

# Get three viridis colors for the nodes
node_colors <- viridis::viridis(3, begin = .25, end = .75)
cloudy_color <- node_colors[1]   # Purple-ish
sunny_color <- node_colors[2]    # Teal-ish
rainy_color <- node_colors[3]    # Yellow-ish

# Edge configuration
edge_resect <- 1
label_size <- 0.45

# Curvature settings for avoiding overlaps
curve_large <- 8.0   # For long distances

# Arc bend settings for inter-node edges
arc_bend_value <- 0.5  # Controls the arc bend of bidirectional edges

# Self-loop offset settings
self_loop_offset <- 20

# ===== CREATE NODES =====
# Define individual node styles with viridis colors
cloudy_style <- ob_style(
  fill = cloudy_color,
  color = "white"
)

sunny_style <- ob_style(
  fill = sunny_color,
  color = "white"
)

rainy_style <- ob_style(
  fill = rainy_color,
  color = "white"
)

large_node_label_3state <- redefault(
  ob_label,
  color = "white",  # White text for better contrast on colored backgrounds
  fill = NA,
  family = my_font,
  size = 18  # Large size for 3-state node labels
)

# Create nodes directly with custom styling
cloudy_node <- ob_circle(
  x = 0, y = 1,
  r = node_radius,
  style = cloudy_style,
  label = large_node_label_3state("Cloudy")
)

sunny_node_3state <- ob_circle(
  x = -1, y = -1,
  r = node_radius,
  style = sunny_style,
  label = large_node_label_3state("Sunny")
)

rainy_node_3state <- ob_circle(
  x = 1, y = -1,
  r = node_radius,
  style = rainy_style,
  label = large_node_label_3state("Rainy")
)

# Position nodes in triangle formation
cloudy_node <- place(cloudy_node, cloudy_node@bounding_box, "top", sep = 20)
sunny_node_3state <-
  place(sunny_node_3state, cloudy_node@bounding_box, "bottom left",
  sep = node_separation)
rainy_node_3state <-
  place(rainy_node_3state, cloudy_node@bounding_box, "bottom right",
  sep = node_separation)

# ===== CREATE EDGE LABELS =====
lbl_ab <- ob_label("0.6",
  angle = 0, position = label_size, size = label_font_size)
lbl_ba <- ob_label("0.1",
  angle = 0, position = label_size, size = label_font_size)
lbl_ac <- ob_label("0.2",
  angle = 0, position = label_size, size = label_font_size)
lbl_ca <- ob_label("0.4",
  angle = 0, position = label_size, size = label_font_size)
lbl_bc <- ob_label("0.2",
  angle = 0, position = label_size, size = label_font_size)
lbl_cb <- ob_label("0.3",
  angle = 0, position = label_size, size = label_font_size)

# Self-loop labels
lbl_aa <- ob_label("?", angle = 0, position = 0.5, size = label_font_size,
                   color = "#f57c7c")
lbl_bb <- ob_label("?", angle = 0, position = 0.5, size = label_font_size,
                   color = "#f57c7c")
lbl_cc <- ob_label("?", angle = 0, position = 0.5, size = label_font_size,
                   color = "#f57c7c")

# ===== CREATE EDGES =====

# ===== SELF-LOOPS =====
# A -> A (self-loop at top with larger offsets)
edge_aa <- connect(
  from = cloudy_node, to = cloudy_node,
  label = lbl_aa, label_sloped = TRUE,
  resect = edge_resect, color = cloudy_color,
  from_offset = ob_point(-self_loop_offset, self_loop_offset * 2.5),
  to_offset = ob_point(self_loop_offset, self_loop_offset * 2.5)
)

# B -> B (self-loop at bottom left with larger offsets)
edge_bb <- connect(
  from = sunny_node_3state, to = sunny_node_3state,
  label = lbl_bb, label_sloped = TRUE,
  resect = edge_resect, color = sunny_color,
  from_offset = ob_point(self_loop_offset, -self_loop_offset * 2.5),
  to_offset = ob_point(-self_loop_offset, -self_loop_offset * 2.5)
)

# C -> C (self-loop at bottom right with larger offsets)
edge_cc <- connect(
  from = rainy_node_3state, to = rainy_node_3state,
  label = lbl_cc, label_sloped = TRUE,
  resect = edge_resect, color = rainy_color,
  from_offset = ob_point(self_loop_offset, -self_loop_offset * 2.5),
  to_offset = ob_point(-self_loop_offset, -self_loop_offset * 2.5)
)

# ===== INTER-NODE EDGES =====
# A -> B (maximum offset + very high curvature)
edge_ab <- connect(
  from = cloudy_node, to = sunny_node_3state,
  label = lbl_ab, label_sloped = FALSE,
  resect = edge_resect, color = cloudy_color,
  curvature = curve_large,
  from_offset = ob_point(-20, -6), to_offset = ob_point(6, 20)
)

# B -> A (maximum different offset + opposite very high curvature)
edge_ba <- connect(
  from = sunny_node_3state, to = cloudy_node,
  label = lbl_ba, label_sloped = FALSE,
  resect = edge_resect, color = sunny_color,
  curvature = -curve_large,
  from_offset = ob_point(20, 6), to_offset = ob_point(-6, -20)
)

# A -> C (symmetric to A -> B with opposite curvature)
edge_ac <- connect(
  from = cloudy_node, to = rainy_node_3state,
  label = lbl_ac, label_sloped = FALSE,
  resect = edge_resect, color = cloudy_color,
  curvature = -curve_large,
  from_offset = ob_point(20, -6), to_offset = ob_point(-6, 20)
)

# C -> A (symmetric to B -> A with opposite curvature)
edge_ca <- connect(
  from = rainy_node_3state, to = cloudy_node,
  label = lbl_ca, label_sloped = FALSE,
  resect = edge_resect, color = rainy_color,
  curvature = curve_large,
  from_offset = ob_point(-20, 6), to_offset = ob_point(6, -20)
)

# B -> C (larger offset for bottom edge)
edge_bc <- connect(
  from = sunny_node_3state, to = rainy_node_3state,
  label = lbl_bc, label_sloped = TRUE,
  resect = edge_resect, color = sunny_color,
  curvature = curve_large,
  from_offset = ob_point(6, 10), to_offset = ob_point(-5, 10)
)

# C -> B (opposite offset to avoid overlap)
edge_cb <- connect(
  from = rainy_node_3state, to = sunny_node_3state,
  label = lbl_cb, label_sloped = TRUE,
  resect = edge_resect, color = rainy_color,
  curvature = -curve_large,
  from_offset = ob_point(-6, -10), to_offset = ob_point(6, -10)
)

# ===== CREATE DIAGRAM =====
diagram <- ggdiagram(font_family = my_font, font_size = font_size) + cloudy_node + sunny_node_3state + rainy_node_3state
diagram <- diagram + edge_aa + edge_bb + edge_cc + edge_ab + edge_ba + edge_ac + edge_ca + edge_bc + edge_cb

# Remove white background
diagram <- diagram + theme_void() + theme(
  panel.background = element_rect(fill = "transparent", color = NA),
  plot.background = element_rect(fill = "transparent", color = NA)
)

print(diagram)
```

Note, the diagram (intentionally) misses out the self-transitions. You should be able to infer this because the probabilities given would otherwise not add up to one!

<button class="button">
  [Show Solution](#mc-solution)
</button>

:::

::: {.infobox}

## Function MC {#mc-function}

Now let's put the code into a function that can take a general transition matrix for any number of states and run the markov chain simulation. The function should be named `run_markov_chain` and take three arguments:

- `transitionMatrix`: a matrix of transition probabilities
- `initialState`: the starting state (an integer)
- `numSteps`: the number of steps to simulate

The function should then return a vector of simulated states as a vector. Once you have written the function, you can test it using the two-state weather model above and the three-state model. Also check that it works for a system with more states. Then also test that the function works for different initial states and number of steps. The final thing to test is if the output is a vector and has the correct length.


<button class="button">
  [Show Solution](#mc-fn-solution)
</button>

:::