# Solution: Representing Data {#representing_data-sol}
## Creating vectors and matrices

```{r}
# Creating matrices
A1 <- matrix(
  c(1:9),
  nrow = 3
)

A2 <- matrix(
  c(1, 3, 4, 6, 9, 2, 1, 0, 3),
  nrow = 3
)

# Creating vectors
v1 <- c(0, 1, 1)
v2 <- c(2, 1, 0)
v3 <- c(3, 1, 1)
x1 <- c(1, 2, 0, 1)
x2 <- c(2, 3, 1, 1)
x3 <- c(4, 1, 2, 0)

# Another way to create matrices
B1 <- cbind(v1, v3, v2)
B2 <- rbind(v1, v2, v3)

```

<button class="button">
  [Back to exercise](#rep_data-ex)
</button>


## Solution Q2

> Q2: Calculate the following matrices.

- `A1 + A2`
- `A1 * A2`
- `A1 - A2`
- transpose of `A1`
- transpose of `B2`

```{r}
A1 + A2

A1 * A2

A1 - A2

t(A1)

t(B2)

```

<button class="button">
  [Back to exercise](#rep_data-ex)
</button>

## Solution Q3

> Q3: Compute the norms of all vectors.

```{r}
sum(v1^2)
sum(v2^2)
sum(v3^2)
sum(x1^2)
sum(x2^2)
sum(x3^2)

```

<button class="button">
  [Back to exercise](#rep_data-ex)
</button>

## Solution Q4

> Q4: Compute the inverse of `A1`, `A2`, and `B3`.
> You don't need to invert the matrices by hand but check they are an inverse.

```{r, error=TRUE}
# Inverse doesn't exist as it is a singular matrix
solve(A1)

solve(A2)

solve(A2) %*% A2

solve(B2)
solve(B2) %*% B2

```

<button class="button">
  [Back to exercise](#rep_data-ex)
</button>

## Solution Q5 and Q6

> Q5: Compute `A1 %*% v1` and `v1 %*% A1`

```{r}
A1 %*% v1

v1 %*% A1
```

> Q6: Compute `v2 %*% B2` and `B2 %*% v2`

```{r}
v2 %*% B2

B2 %*% v2
```

<button class="button">
  [Back to exercise](#rep_data-ex)
</button>