# (APPENDIX) Solutions {-}

# Solution: Simulations {#simSol}

## Solution Simulation - replacement {#sim-sol-ex1}

The option `replace=TRUE` activates sampling with replacement (i.e. the numbers that are picked are put back and can be picked again).

The option `replace=FALSE` activates sampling without replacement (i.e. the numbers that are picked are not put back and cannot be picked again).

Let's try this out:

```{r, error = TRUE}
x <- c( 1, 2, 2, 3, 4, 1, 6, 7, 8, 10, 5, 5, 1, 4, 9 )

# Working example
sample(x, 10, replace = FALSE)

# This will cause an error
sample(x, 20, replace = FALSE)

# Fix it
sample(x, 20, replace = TRUE)
```

<button class="button">
  [Back to exercise](#sim-ex1)
</button>

## Solution Simulation - using sample {#sim-sol-ex2}


```{r}
rolls_from_sample = sample(c(1:6), size=5000, replace=TRUE)
rolls_from_sample.int = sample(6, size=5000, replace=TRUE)

table(rolls_from_sample)
```

```{r}
table(rolls_from_sample.int)
```

Both gives a uniform distribution over the numbers 1-6. The function `sample.int` is a specialised version of sample for sampling integers. Many `R` libraries have specialised versions of more general functions to do specific tasks under certain conditions.

<button class="button">
  [Back to exercise](#sim-ex2)
</button>
